California Road Construction Safety Analysis: A Spatial Econometric Approach

### 2025-03-26 final/final_MoritzPeistNoemiLucchiSimonVellin.Rmd
Restructured, cleaned, added final code

---
title: "R Notebook"
output: html_notebook
---

Load packages

```{r Packages}
library(data.table)   # For faster data manipulation
library(tidyverse)    # For data manipulation and visualization
library(sf)           # For spatial data handling
library(leaflet)      # For interactive maps
library(leaflet.extras) # For additional leaflet features
library(mapview)      # For easier map visualization
library(tmap)         # For thematic maps
library(tigris)       # For US road networks
library(future)       # For parallel processing
library(future.apply) # For parallel processing with apply functions

@@

# Set custom cache directory (optional)
options(tigris_cache_dir = "./data/tigris")
# Configure tigris to use caching
options(tigris_use_cache = TRUE)
```

# Set-up convenience functions for parallel processing (as data is quite large)
```{r Parallel Buffering}
# Function to create buffers in batches with proper projection
create_buffers_in_batches <- function(sf_object, buffer_dist, batch_size = 500) {
  # First, reproject to an appropriate projected CRS for the region
  # For California, UTM Zone 10N (EPSG:26910) or 11N (EPSG:26911) works well
  # If we're unsure about your region, a web mercator projection (3857) is a reasonable default
  message("Reprojecting data to a meter-based CRS...")
  sf_object_projected <- st_transform(sf_object, 3310)  # Web Mercator

  n_features <- nrow(sf_object_projected)
  n_batches <- ceiling(n_features / batch_size)

  # Create empty list to store results
  buffer_list <- vector("list", n_batches)

  # Process in batches
  for (i in 1:n_batches) {
    start_idx <- (i-1) * batch_size + 1
    end_idx <- min(i * batch_size, n_features)

    cat(sprintf("Processing batch %d of %d (features %d to %d)\n", 
                i, n_batches, start_idx, end_idx))

    # Extract batch
    batch <- sf_object_projected[start_idx:end_idx, ]

    # Create buffer (with parallel processing within sf)
    buffer_list[[i]] <- st_buffer(batch, dist = buffer_dist)
  }

  # Combine results
  result <- do.call(rbind, buffer_list)

  # Reproject back to original CRS if needed
  message("Reprojecting results back to original CRS...")
  st_transform(result, st_crs(sf_object))
}
```

Load data (filter for California for now to limit data set size)

```{r Data Load Accidents}
# Efficient approach
df.acc <- fread("data/us_accidents/US_accidents_March23.csv")[
  # Filter date range to 2016-2021
  as.Date(Start_Time) >= as.Date("2016-01-01") & 
  as.Date(Start_Time) <= as.Date("2021-12-31") & 
  # And California
  State == "CA"
][, `:=`(
  # Add year, quarter, month columns
  year = data.table::year(Start_Time),
  quarter = data.table::quarter(Start_Time),
  month = data.table::month(Start_Time),
  # Calculate duration (assuming End_Time exists in the dataset)
  duration = as.numeric(difftime(End_Time, Start_Time, units = "days"))
)] %>% 
  as_tibble()  # Convert to tibble only at the end for performance
```

```{r Data Load Constructions}
df.const <- fread("data/us_constructions/US_constructions_Dec21.csv")[
  # Filter date range to 2016-2021
  as.Date(Start_Time) >= as.Date("2016-01-01") & 
  as.Date(Start_Time) <= as.Date("2021-12-31") & 
    # And California
  State == "CA"
][, `:=`(
  # Add year, quarter, month columns
  year = year(Start_Time),
  quarter = quarter(Start_Time),
  month = month(Start_Time),
  # Calculate duration (assuming End_Time exists in the dataset)
  duration = as.numeric(difftime(End_Time, Start_Time, units = "days"))
)] %>% 
  as_tibble()  # Convert to tibble only at the end for performance
```

Convert dataframes to sf objects

```{r Geospatial Data}
# Convert accident data to sf object
df.acc.sf <- df.acc %>%
  filter(!is.na(Start_Lat) & !is.na(Start_Lng)) %>%
  st_as_sf(coords = c("Start_Lng", "Start_Lat"), crs = 4326)

# Convert construction data to sf object
df.const.sf <- df.const %>%
  filter(!is.na(Start_Lat) & !is.na(Start_Lng)) %>%
  st_as_sf(coords = c("Start_Lng", "Start_Lat"), crs = 4326)
```

```{r Data Load Roads}
# Get US roads (this can be slow for the entire US, so we might want to limit by state)

# First get all counties in California
ca_counties <- counties("CA", year = 2021)

# Then get roads for each county
ca_roads_list <- lapply(ca_counties$COUNTYFP, function(county_code) {
  roads("CA", county = county_code, year = 2021)
})

# Optionally combine all county road data into one object
ca_roads <- do.call(rbind, ca_roads_list)
```

```{r Static Maps}
# Static map for Accidents with year coloring
# First, store your plot in a variable
ca_plot <- ggplot() +
  geom_sf(data = ca_roads, color = "gray80", size = 0.1) +
  geom_sf(data = df.acc.sf %>% filter(State == "CA"), 
          aes(color = factor(year)), alpha = 0.7, size = 0.5) +
  scale_color_viridis_d(name = "Year") +  # Use viridis color palette for discrete values
  theme_minimal() +
  labs(title = "US Road Accidents by Year (2016-2021)") +
  theme(legend.position = "bottom")
print(ca_plot)
# Then save it using ggsave
#ggsave(filename = "imgs/california_accidents.png", 
#       plot = ca_plot,
#       width = 10, # width in inches
#       height = 8, # height in inches
#       dpi = 300)  # resolution

# Construction sites map with year coloring
# First, store our plot in a variable
ca_const_plot <- ggplot() +
  geom_sf(data = ca_roads, color = "gray80", size = 0.1) +
  geom_sf(data = df.const.sf %>% filter(State == "CA"), 
          aes(color = factor(year)), alpha = 0.7, size = 0.5) +
  scale_color_viridis_d(name = "Year") +  # Use viridis color palette for discrete values
  theme_minimal() +
  labs(title = "US Construction Sites by Year (2016-2021)") +
  theme(legend.position = "bottom")
print(ca_const_plot)
# Then save it using ggsave
#ggsave(filename = "imgs/california_construction.png", 
#       plot = ca_const_plot,
#       width = 10, # width in inches
#       height = 8, # height in inches
#       dpi = 300)  # resolution
```

## Heatmaps - interactive

```{r Interactive Maps I}
# For interactive maps (often better for large datasets)
# Accidents map
accident_map <- leaflet() %>%
  addTiles() %>%
  addHeatmap(data = df.acc.sf, 
             intensity = ~1,
             radius = 8, 
             blur = 10) %>%
  setView(lng = -119.4179, lat = 36.7783, zoom = 6) %>% # Center to CA
  setMaxBounds(lng1 = -124.6, lat1 = 42.0,    # Northwest corner of CA
               lng2 = -114.1, lat2 = 32.5)    # Southeast corner of CA

# Display maps
accident_map
```

```{r Interactive Maps II}
# Construction sites map with California focus and bounds constraints
construction_map <- leaflet() %>%
  addTiles() %>%
  addHeatmap(data = df.const.sf, 
             intensity = ~1,
             radius = 8, 
             blur = 10,
             gradient = c("yellow", "orange", "red")) %>%
  setView(lng = -119.4179, lat = 36.7783, zoom = 6) %>%  # Center on California
  setMaxBounds(lng1 = -124.6, lat1 = 42.0,    # Northwest corner of CA
               lng2 = -114.1, lat2 = 32.5)    # Southeast corner of CA

# Display the map
construction_map
```

## Buffer zone analysis

```{r Buffer Zone Analysis}
# Set up parallel processing
future::plan(future::multisession, workers = parallel::detectCores() - 1)

# Set sf to use parallel processing
sf_use_s2(FALSE)  # Disable S2 spherical geometry
cores <- parallel::detectCores() - 1

# Process each dataset with batching
cat("Creating point buffers...\n")
const_buffers <- create_buffers_in_batches(df.const.sf, buffer_dist = 1000, batch_size = 500)

# Return to sequential processing
future::plan(future::sequential)
sf_use_s2(TRUE)  # Re-enable S2 spherical geometry

# Find accidents within construction zones
# Use a spatial index and parallel processing
plan(multisession, workers = parallel::detectCores() - 1)  # Use parallel processing

accidents_in_construction <- df.acc.sf %>%
  st_join(ca_const_buffers, join = st_within, prepared = TRUE)

# Reset to sequential processing when done
plan(sequential)

# Count accidents per construction zone
accidents_per_zone <- accidents_in_construction %>%
  group_by(ID) %>%  # Grouping by the ID column
  summarize(accident_count = n())

# Visualize construction zones with accident counts
# First, store our plot in a variable
accident_zones_plot <- ggplot() +
  geom_sf(data = ca_roads, color = "gray80", size = 0.1) +
  geom_sf(data = accidents_per_zone, aes(fill = accident_count), alpha = 0.7) +
  scale_fill_viridis_c() +
  theme_minimal() +
  labs(title = "Accident Counts within Construction Zones (2016-2021)")

print(accident_zones_plot)

# Then save it using ggsave
#ggsave(filename = "imgs/accident_counts_construction_zones.png", 
#       plot = accident_zones_plot,
#       width = 10, # width in inches
#       height = 8, # height in inches
#       dpi = 300)  # resolution
```

Advanced construction sites:

```{r SF linestring}
# Convert construction data to linestring sf object
df.const.lines <- df.const %>%
  filter(!is.na(Start_Lat) & !is.na(Start_Lng) & !is.na(End_Lat) & !is.na(End_Lng)) %>%
  mutate(geometry = pmap(list(Start_Lng, Start_Lat, End_Lng, End_Lat), 
                         function(start_lng, start_lat, end_lng, end_lat) {
                           coords <- matrix(c(start_lng, start_lat, 
                                             end_lng, end_lat), 
                                           ncol = 2, byrow = TRUE)
                           st_linestring(coords)
                         })) %>%
  st_sf(crs = 4326)
```

```{r Maps with Construction Lines}
# Static map with construction line segments
ggplot() +
  geom_sf(data = ca_roads, color = "gray80", size = 0.1) +
  geom_sf(data = df.const.lines %>% filter(State == "CA"), 
          color = "orange", size = 1) +
  theme_minimal() +
  labs(title = "US Construction Segments (2016-2021)")

# Interactive map with construction segments focused on California
construction_segment_map <- leaflet() %>%
  addTiles() %>%
  addPolylines(data = df.const.lines, 
               color = "orange",
               weight = 3,
               opacity = 0.7,
               popup = ~paste("ID:", ID, "<br>Duration:", Duration)) %>%
  setView(lng = -119.4179, lat = 36.7783, zoom = 6) %>%  # Center on California
  setMaxBounds(lng1 = -124.6, lat1 = 42.0,    # Northwest corner of CA
               lng2 = -114.1, lat2 = 32.5)    # Southeast corner of CA

# Display the map
construction_segment_map
```

```{r Construction Impact Analysis with Buffers}
# Set up parallel processing
future::plan(future::multisession, workers = parallel::detectCores() - 1)

# Set sf to use parallel processing
sf_use_s2(FALSE)  # Disable S2 spherical geometry
cores <- parallel::detectCores() - 1

cat("Creating line segment buffers...\n")
const_segment_buffers <- create_buffers_in_batches(df.const.lines, buffer_dist = 1000, batch_size = 250)

# Return to sequential processing
future::plan(future::sequential)
sf_use_s2(TRUE)  # Re-enable S2 spherical geometry

# Find accidents near construction segments
accidents_near_construction <- df.acc.sf %>%
  st_join(const_segment_buffers, join = st_intersects)

# Analyze accident rate per construction segment length
accidents_per_segment <- accidents_near_construction %>%
  group_by(ID) %>%  # Assuming there's an ID column for construction segments
  summarize(
    accident_count = n(),
    segment_length = as.numeric(st_length(geometry[1])),  # Length in meters
    accidents_per_km = accident_count / (segment_length/1000)
  )

# Visualize segments by accident rate
ggplot() +
  geom_sf(data = ca_roads, color = "gray80", size = 0.1) +
  geom_sf(data = accidents_per_segment, aes(color = accidents_per_km), size = 1.5) +
  scale_color_viridis_c() +
  theme_minimal() +
  labs(title = "Accident Rate per Construction Segment (2016-2021)",
       color = "Accidents/km")
```

```{r Temporal Analysis and Severity Comparison with Buffers}
# Temporal analysis - construction impact over time
accidents_by_time <- accidents_near_construction %>%
  mutate(month = floor_date(as.Date(Start_Time), "month")) %>%
  group_by(month) %>%
  summarize(accident_count = n())

# Compare accident severity near vs. away from construction
severity_comparison <- bind_rows(
  accidents_near_construction %>% 
    mutate(location = "Near construction") %>% 
    select(Severity, location),

  anti_join(df.acc.sf, accidents_near_construction, by = "ID") %>% 
    mutate(location = "Away from construction") %>% 
    select(Severity, location)
)

# Visualize severity comparison
ggplot(severity_comparison, aes(x = Severity, fill = location)) +
  geom_bar(position = "dodge") +
  scale_fill_brewer(palette = "Set1") +
  theme_minimal() +
  labs(title = "Accident Severity: Near vs. Away from Construction",
       x = "Severity Level",
       y = "Count",
       fill = "Location")
```

>>>>

---
title: "California Road Construction Safety Analysis: A Spatial Econometric Approach"
author:
- Moritz Peist
- Noemi Lucchi
- Simon Vellin
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    df_print: paged
    code_folding: show
    fig_caption: true
    number_sections: true
    progress: false
    warning: false
---

```{css, echo=FALSE}
h1.title {
  text-align: center;
}

h4.author, h4.date {
  text-align: center;
}

.author {
  text-align: center;
}

.date {
  text-align: center;
}

h3.subtitle {
  text-align: center;
  font-style: italic;
  margin-top: -10px;
  color: #606060;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE, 
  message = FALSE,
  fig.width = 8,
  fig.height = 5,
  out.width = "100%",
  fig.align = "center",
  comment = "#>"
)

# Load required libraries
library(data.table)      # For faster data manipulation
library(tidyverse)       # For data manipulation and visualization
library(sf)              # For spatial data handling
library(leaflet)         # For interactive maps
library(leaflet.extras)  # For additional leaflet features
library(mapview)         # For easier map visualization
library(tmap)            # For thematic maps
library(tigris)          # For US road networks
library(fixest)          # For fixed effects models
library(spdep)           # For spatial weights and tests
library(splm)            # For spatial panel models
library(RColorBrewer)    # For color palettes
library(patchwork)       # For combining plots
library(kableExtra)      # For nice tables
library(lubridate)       # For date manipulation
library(viridis)         # For color palettes

@@

# Set custom cache directory
options(tigris_cache_dir = "./data/tigris")
# Configure tigris to use caching
options(tigris_use_cache = TRUE)
```

<div style="text-align: center;">
  <img src="../Barcelona_School_of_Economics_logo.svg" alt="BSE logo" style="width: 200px;">
</div>

# Abstract

This study examines the causal relationship between road construction activities and traffic accident rates in California using high-resolution spatial data from 2021. Implementing a difference-in-differences framework with spatial controls, we analyze how construction zones affect accident frequency and severity. Our findings indicate a significant relationship between active construction and accident occurrence, with heterogeneous effects across urban and rural environments. Spatial spillover analysis reveals that accident impacts extend beyond immediate construction zones, with pronounced effects in areas surrounding construction sites. This research provides evidence-based insights for transportation safety planning and targeted mitigation strategies.

# Introduction

Road construction is a necessary yet potentially disruptive component of transportation infrastructure maintenance and development. While essential for long-term safety improvements, construction activities create temporary changes in traffic patterns, road conditions, and driver behavior that may increase accident risk. Understanding the causal relationship between construction activities and traffic safety has important implications for construction scheduling, resource allocation, traffic management strategies, and policy development.

Despite the importance of this relationship, rigorous causal evidence on how road construction affects traffic accidents remains limited, particularly with respect to potential heterogeneity between urban and rural environments. This study addresses this gap by leveraging comprehensive spatial data on road construction activities and traffic accidents in California during 2021 to answer the following research questions:

1. How does road construction causally impact traffic accident rates?
2. Do these effects systematically differ between urban and rural environments?
3. Are there spatial spillover effects beyond the immediate construction zones?

The findings of this research have direct implications for transportation safety planning, particularly regarding how resources should be allocated for traffic management during construction activities across different environments.

# Literature Review

To establish a contextual foundation for our study, we reviewed existing literature on the relationship between road construction and traffic accidents, focusing on studies that use spatial or causal inference approaches.

## 2.1 ["Effects of truck traffic on crash injury severity on rural highways in Wyoming using Bayesian binary logit models" by Ahmed et al. (2018)](https://www.sciencedirect.com/science/article/pii/S0001457518301507)

This paper explores the impact of truck traffic and roadway geometry on crash frequency within highway work zones, using a case study from Wyoming spanning 2012–2016.

Using crash data and statistical modeling, the authors assess how construction-related factors, like lane closures and heavy vehicle presence, elevate accident risks. Their paper reveals that increased truck volumes and sharper geometric features significantly amplify crash rates, with work zones acting as hotspots for such incidents. While the paper employs logistic regression rather than causal inference techniques, it underscores the real-world relevance of construction disruptions on traffic safety.

This work provides a practical foundation for our study, highlighting the need to isolate causal effects of construction using advanced methods. By focusing on observable crash patterns, Ahmed et al. set the stage for deeper methodological exploration in our difference-in-differences framework.

## 2.2 ["Comparison of characteristics between fatal and injury accidents in the highway construction zones" by Li and Bai (2008)]( https://www.sciencedirect.com/science/article/pii/S0925753507001014)

This paper investigates the characteristics of accidents in highway construction zones from a statistical perspective, comparing fatal and injury crashes using data from Utah between 1992 and 2004.

Using traffic volume, construction duration, and time of day, their results show that construction zones markedly increase accident likelihood, with fatal crashes tied to higher speeds and nighttime conditions. Although the study relies on descriptive statistics and regression rather than causal frameworks, it offers critical insights into how construction alters road safety dynamics.

## Accident Risk in Highway Construction Zones

The relationship between construction activities and traffic accidents has been investigated in several studies. Ahmed et al. (2018) explored the impact of truck traffic and roadway geometry on crash frequency within highway work zones in Wyoming. Their analysis revealed that increased truck volumes and sharper geometric features significantly amplified crash rates, with work zones acting as hotspots for accidents. While their study employed logistic regression rather than causal inference techniques, it highlighted the real-world relevance of construction disruptions on traffic safety.

Li and Bai (2008) compared the characteristics of fatal and injury crashes in highway construction zones using data from Utah (1992-2004). Their results showed that construction zones markedly increase accident likelihood, with fatal crashes tied to higher speeds and nighttime conditions. Although they relied on descriptive statistics rather than causal frameworks, their work established a clear link between construction activities and accident rates.

The literature generally suggests that construction zones are associated with increased accident risk, but most studies have not employed causal inference methods to isolate this effect from confounding factors. Additionally, few studies have examined the spatial heterogeneity of construction effects on accidents, particularly regarding urban-rural differences. Our study addresses these gaps by employing a difference-in-differences framework with spatial controls to identify the causal impact of construction activities on accident rates.

# Data and Methods

## Data Sources

This study integrates multiple large-scale datasets to create a comprehensive spatial-temporal framework for analysis:

**Primary Datasets:**

  1. [**US Road Construction and Closures Dataset (2021)**](https://www.kaggle.com/datasets/sobhanmoosavi/us-road-construction-and-closures) - Contains detailed records of construction activities in California with coordinates, start/end times, and affected distances
  2. [**US Traffic Accidents Dataset (2021)**](https://www.kaggle.com/datasets/sobhanmoosavi/us-accidents) - Comprehensive accident records with precise locations, timestamps, and severity measures
  3. [**Caltrans State Highway Network**](https://catalog.data.gov/dataset/traffic-volumes-aadt-ee8d6/resource/904a8300-9b57-4765-8ed7-cca0c83cdede) - Annual Average Daily Traffic (AADT) data for normalizing accident rates by traffic volume
  4. **US Census Bureau Urban Areas** - Official urban/rural classifications for spatial heterogeneity analysis

For both the accident and construction datasets, the data was collected using multiple APIs such as the US and state departments of transportation, law enforcement agencies, traffic cameras, and traffic sensors within the road networks.

While our analysis uses 2021 accident and construction data, we integrated Annual Average Daily Traffic (AADT) data from Caltrans to normalize accident rates. The AADT dataset provides a comprehensive traffic volume baseline that reflects typical traffic patterns, mitigating potential short-term fluctuations. Though the AADT data may include projections or estimates from surrounding years, it represents the most reliable and granular traffic volume information available, enabling more precise exposure-adjusted accident rate calculations.

## Data Preparation and Exploration

We focused our analysis on California road construction and accidents from 2021, creating a spatial-temporal framework that allows us to examine the causal relationship between construction activities and accidents.

```{r data_loading}
# Load Caltrans State Highway Network
aadt <- st_read("data/Traffic_Volumes_AADT/Traffic_Volumes_AADT.shp")

# Load and filter accidents data
df.acc <- fread("data/us_accidents/US_accidents_March23.csv")[
  # Filter date range of 2021
  lubridate::year(as.Date(Start_Time)) == 2021 & 
  # And California
  State == "CA"
][, `:=`(
  # Add year, quarter, month columns
  year = data.table::year(Start_Time),
  quarter = data.table::quarter(Start_Time),
  month = data.table::month(Start_Time),
  # Calculate duration
  duration = as.numeric(difftime(End_Time, Start_Time, units = "days"))
)] %>% 
  as_tibble()

# Load and filter construction data
df.const <- fread("data/us_constructions/US_constructions_Dec21.csv")[
  # Filter date range of 2021
  lubridate::year(as.Date(Start_Time)) == 2021 & 
  # And California
  State == "CA"
][, `:=`(
  # Add year, quarter, month columns
  year = year(Start_Time),
  quarter = quarter(Start_Time),
  month = month(Start_Time),
  # Calculate duration
  duration = as.numeric(difftime(End_Time, Start_Time, units = "days"))
)][
  duration > 1  # Filter out constructions lasting less than a day
] %>% 
  as_tibble()

# Load California road network for mapping
ca_roads <- tigris::primary_roads("CA", year = 2021, refresh=TRUE) %>%
  st_transform(4326)
```

```{r data_summary}
# Calculate and display actual data summary
accidents_count <- nrow(df.acc)
construction_count <- nrow(df.const)
aadt_count <- nrow(aadt)

cat("Summary of datasets:\n")
cat("- Accidents dataset:", accidents_count, "records\n")
cat("- Construction dataset:", construction_count, "records\n")
cat("- AADT dataset:", aadt_count, "records\n")
```

### Temporal Patterns

```{r temporal_patterns}
# Temporal patterns visualization using actual data
acc_monthly <- df.acc %>%
  mutate(month = floor_date(as.Date(Start_Time), "month")) %>%
  count(month)

const_monthly <- df.const %>%
  mutate(month = floor_date(as.Date(Start_Time), "month")) %>%
  count(month)

# Plot monthly patterns
p1 <- ggplot(acc_monthly, aes(x = month, y = n)) +
  geom_line(color = "blue", size = 1) +
  geom_point(color = "blue", size = 3) +
  labs(title = "Monthly Accidents in California (2021)",
       x = "Month", y = "Count") +
  scale_x_date(date_labels = "%b", date_breaks = "1 month") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

p2 <- ggplot(const_monthly, aes(x = month, y = n)) +
  geom_line(color = "red", size = 1) +
  geom_point(color = "red", size = 3) +
  labs(title = "New Construction Projects in California (2021)",
       x = "Month", y = "Count") +
  scale_x_date(date_labels = "%b", date_breaks = "1 month") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

p1 / p2
```

The temporal distribution reveals distinct patterns for accidents and construction projects. Accident rates peak in winter months (January and December) and fall (October), while construction activity is highest in spring (March-May). This inverse relationship suggests seasonal factors may influence both construction planning and accident occurrence.

### Spatial Distribution

Let's examine the spatial distribution of accidents and construction sites across California.

```{r spatial_processing}
# Create spatial objects from actual data
if (!inherits(df.acc, "sf")) {
  df.acc.sf <- df.acc %>%
    filter(!is.na(Start_Lat) & !is.na(Start_Lng)) %>%
    st_as_sf(coords = c("Start_Lng", "Start_Lat"), crs = 4326) %>%
    st_transform(crs = 3310)  # CA Albers for accurate distance
} else {
  df.acc.sf <- df.acc
}

if (!inherits(df.const, "sf")) {
  df.const.sf <- df.const %>%
    filter(!is.na(Start_Lat) & !is.na(Start_Lng)) %>%
    st_as_sf(coords = c("Start_Lng", "Start_Lat"), crs = 4326) %>%
    st_transform(crs = 3310)
} else {
  df.const.sf <- df.const
}
```

```{r static_maps}
# Create a subsample for visualization (plotting all points would be too dense)
set.seed(123)
acc_sample <- df.acc.sf %>%
  sample_n(min(10000, nrow(.)))

const_sample <- df.const.sf %>%
  sample_n(min(5000, nrow(.)))

# Static map for Accidents with month coloring
ca_acc_plot <- ggplot() +
  geom_sf(data = ca_roads, color = "gray80", size = 0.1) +
  geom_sf(data = acc_sample, 
          aes(color = factor(month)), alpha = 0.5, size = 0.5) +
  scale_color_viridis_d(name = "Month", option = "plasma") +
  coord_sf(
    xlim = c(-124.5, -114.5),
    ylim = c(32.5, 42.5)
  ) +
  theme_minimal() +
  labs(title = "California Road Accidents (2021)") +
  theme(legend.position = "bottom")

# Construction sites map with severity coloring
ca_const_plot <- ggplot() +
  geom_sf(data = ca_roads, color = "gray80", size = 0.1) +
  geom_sf(data = const_sample, 
          aes(color = factor(Severity)), alpha = 0.7, size = 0.5) +
  scale_color_viridis_d(name = "Severity", option = "plasma") +
  coord_sf(
    xlim = c(-124.5, -114.5),
    ylim = c(32.5, 42.5)
  ) +
  theme_minimal() +
  labs(title = "California Construction Sites by Severity (2021)") +
  theme(legend.position = "bottom")

# Display the plots
ca_acc_plot
ca_const_plot
```

### Density Heatmaps

To better visualize the spatial concentration, we created heatmaps for both accidents and construction sites.

```{r heatmaps}
# Sample data for creating heatmaps (using smaller sample for performance)
set.seed(123)
acc_sample_for_heatmap <- df.acc %>%
  filter(!is.na(Start_Lat) & !is.na(Start_Lng)) %>%
  sample_n(min(50000, nrow(.)))

const_sample_for_heatmap <- df.const %>%
  filter(!is.na(Start_Lat) & !is.na(Start_Lng)) %>%
  sample_n(min(10000, nrow(.)))

# Create heatmap of accidents using ggplot2
accident_heatmap <- ggplot() +
  geom_sf(data = ca_roads, color = "gray80", size = 0.1) +
  stat_density_2d(
    data = acc_sample_for_heatmap,
    aes(x = Start_Lng, y = Start_Lat, fill = after_stat(density), alpha = after_stat(density)),
    geom = "tile", 
    contour = FALSE,
    h = 0.2,  # Increased bandwidth for better visualization
    n = 150
  ) +
  scale_alpha_continuous(range = c(0, 0.9), guide = "none") +
  scale_fill_gradientn(
    colors = c("#0000FF", "#00FFFF", "#00FF00", "#FFFF00", "#FF0000"),
    name = "Accident\nDensity"
  ) +
  coord_sf(
    xlim = c(-124.5, -114.5),
    ylim = c(32.5, 42.5)
  ) +
  theme_minimal() +
  labs(
    title = "Accident Density Heatmap in California (2021)",
    x = NULL,
    y = NULL
  ) +
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold")
  )

# Construction sites heatmap
construction_heatmap <- ggplot() +
  geom_sf(data = ca_roads, color = "gray80", size = 0.1) +
  stat_density_2d(
    data = const_sample_for_heatmap,
    aes(x = Start_Lng, y = Start_Lat, fill = after_stat(density)),
    geom = "tile", 
    contour = FALSE,
    alpha = 0.7,
    h = 0.2,
    n = 150
  ) +
  scale_fill_gradientn(
    colors = c("yellow", "orange", "red"),
    name = "Construction\nDensity",
    na.value = NA,  # Makes low-density areas transparent
    limits = c(0.001, NA)  # Sets minimum value to show (adjust as needed)
  ) +
  coord_sf(
    xlim = c(-124.5, -114.5),
    ylim = c(32.5, 42.5)
  ) +
  theme_minimal() +
  labs(
    title = "Construction Density Heatmap in California (2021)",
    x = NULL,
    y = NULL
  ) +
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold")
  )

# Display the heatmaps
accident_heatmap
construction_heatmap
```

The heatmaps confirm the spatial concentration of both accidents and construction activities in California's major urban areas. The similar patterns suggest a potential relationship, but we need to control for traffic volume differences and establish causality.

### Urban-Rural Classification

We classified construction sites as urban or rural based on their intersection with US Census Bureau urban areas. This classification is crucial for our heterogeneity analysis.

```{r urban_rural}
# Download urban areas for California
ca_urban_areas <- tigris::urban_areas(cb = TRUE, year = 2019) %>%
  st_transform(st_crs(df.const.sf)) %>%
  filter(grepl(", CA$", NAME10))

# Classify construction sites as urban or rural
const_urban <- st_join(
  df.const.sf,
  ca_urban_areas %>% select(urban_name = NAME10),
  join = st_intersects
)

const_urban <- const_urban %>%
  mutate(is_urban = !is.na(urban_name))
```

```{r urban_rural_table}
# Calculate actual urban-rural distribution
urban_rural_df <- const_urban %>%
  st_drop_geometry() %>%
  count(is_urban) %>%
  mutate(percentage = n / sum(n) * 100)

kableExtra::kable(urban_rural_df, 
                 col.names = c("Urban Area", "Count", "Percentage (%)"),
                 caption = "Distribution of Construction Sites by Urban/Rural Classification",
                 booktabs = TRUE,
                 digits = 1) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                           full_width = FALSE) %>%
  kableExtra::row_spec(0, bold = TRUE)
```

Our analysis of the urban-rural breakdown is critical for understanding the heterogeneous effects of construction on accident rates.

### Integration with Traffic Volume Data

To control for exposure to traffic, we integrated Annual Average Daily Traffic (AADT) data from Caltrans, allowing us to normalize accident counts by traffic volume.

```{r traffic_volume}
# Transform AADT data to ensure matching CRS
aadt_sf <- st_transform(aadt, st_crs(df.const.sf))

# Convert AADT columns to numeric
aadt_sf <- aadt_sf %>%
  mutate(
    BACK_AADT_num = as.numeric(as.character(BACK_AADT)),
    AHEAD_AADT_num = as.numeric(as.character(AHEAD_AADT))
  )

# Find the nearest AADT point for each construction buffer
nearest_indices <- st_nearest_feature(df.const.sf, aadt_sf)
nearest_distances <- st_distance(df.const.sf, aadt_sf[nearest_indices,], by_element = TRUE)

# Create a dataframe with buffer IDs and nearest AADT values
buffer_aadt <- df.const.sf %>%
  mutate(buffer_id = paste0("c", row_number())) %>%
  select(buffer_id) %>%
  bind_cols(
    aadt_sf[nearest_indices, ] %>%
    select(BACK_AADT_num, AHEAD_AADT_num)
  ) %>%
  mutate(
    distance_to_nearest = as.numeric(nearest_distances),
    traffic_volume = rowMeans(cbind(BACK_AADT_num, AHEAD_AADT_num), na.rm = TRUE)
  ) %>%
  st_drop_geometry() %>%
  select(buffer_id, traffic_volume, distance_to_nearest)

# Summary of traffic volumes
summary(buffer_aadt$traffic_volume)
```

```{r traffic_distribution}
# Plot actual traffic volume distribution
ggplot(buffer_aadt, aes(x = traffic_volume)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "black", alpha = 0.8) +
  labs(
    title = "Distribution of Annual Average Daily Traffic (AADT)",
    x = "Traffic Volume (vehicles per day)",
    y = "Count"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

# Log-scale for better visualization of the distribution
ggplot(buffer_aadt, aes(x = traffic_volume)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "black", alpha = 0.8) +
  scale_x_log10() +
  labs(
    title = "Distribution of Annual Average Daily Traffic (AADT) - Log Scale",
    x = "Traffic Volume (vehicles per day) - Log Scale",
    y = "Count"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

The traffic volume distribution shows a pattern likely reflecting the differences between urban and rural areas. Urban areas typically have higher traffic volumes, which could influence both construction planning and accident rates.

## Construction Buffer Analysis

To establish the spatial relationship between accidents and construction activities, we created buffer zones around construction locations and analyzed accidents within these zones.

```{r construction_buffers}
# Create construction buffers (500m)
const_buffers <- const_urban %>%
  mutate(buffer_id = paste0("c", row_number())) %>%
  st_buffer(dist = 500)  # 500m buffer

# Add buffer metadata
const_buffers <- const_buffers %>%
  mutate(
    start_date = as.Date(Start_Time),
    end_date = as.Date(End_Time),
    construction_duration = as.numeric(difftime(end_date, start_date, units = "days"))
  )

# Join accidents to construction buffers - this operation can be computationally intensive
# For very large datasets, consider using a spatial database or partitioning the data
accidents_in_buffers <- st_join(
  df.acc.sf %>% 
    select(accident_id = ID, accident_date = Start_Time, Severity),
  const_buffers %>% 
    select(buffer_id, construction_id = ID, start_date, end_date, is_urban),
  join = st_intersects
)

# Classify accidents by timing relative to construction
accidents_in_buffers <- accidents_in_buffers %>%
  mutate(
    accident_date = as.Date(accident_date),
    time_period = case_when(
      is.na(start_date) ~ NA_character_,
      accident_date < start_date ~ "before",
      accident_date > end_date ~ "after",
      TRUE ~ "during"
    ),
    month_year = floor_date(accident_date, "month")
  )

# Display the first few rows of the joined data
head(accidents_in_buffers %>% st_drop_geometry())
```

```{r period_counts_table}
# Count accidents by construction period from actual data
period_df <- accidents_in_buffers %>%
  st_drop_geometry() %>%
  filter(!is.na(time_period)) %>%
  count(time_period) %>%
  mutate(percentage = n / sum(n) * 100)

kableExtra::kable(period_df, 
                 col.names = c("Construction Period", "Count", "Percentage (%)"),
                 caption = "Distribution of Accidents by Construction Period",
                 booktabs = TRUE,
                 digits = 1) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"),
                           full_width = FALSE) %>%
  kableExtra::row_spec(0, bold = TRUE)
```

When examining accidents by their timing relative to construction, this distribution provides initial evidence that construction activities may impact accident patterns, but we need more rigorous analysis to establish causality.

```{r buffer_visualization}
# Extract LA area data for visualization
la_bbox <- st_bbox(c(xmin = -118.75, ymin = 33.65, xmax = -117.85, ymax = 34.35), crs = st_crs(4326))
la_bbox_sf <- st_as_sfc(la_bbox)

# Filter construction buffers and accidents for LA area
la_buffers <- const_buffers %>%
  st_transform(4326) %>%
  st_filter(la_bbox_sf)
